---
title: "Assignment04"
author: "Xiang Xiang and Weijing Zhong"
format: html
---

### Data Source
[IPUMS CURRENT POPULATION SURVEY](https://cps.ipums.org/cps/index.shtml).

```{r}

library(ggplot2)
library(tidyverse)
library(dplyr)
library(ipumsr)

ddi <- read_ipums_ddi("cps_00003.xml")
data <- read_ipums_micro(ddi)

glimpse(data)

```


### Plot 1 Employment Status by Gender

```{r}

cps_2024 <- data %>%
  filter(YEAR == 2024, !EMPSTAT %in% c(0, 1))

cps_2024$SEX <- factor(cps_2024$SEX, levels = c(1, 2), labels = c("Male", "Female"))

cps_2024$EMP_GROUP <- ifelse(cps_2024$EMPSTAT %in% c(10, 12), "Employed",
                             ifelse(cps_2024$EMPSTAT %in% c(20, 21, 22), "Unemployed", "NILF"))

cps_2024$EMP_GROUP <- factor(cps_2024$EMP_GROUP, levels = c("Employed", "Unemployed", "NILF"))

ggplot(data = cps_2024, aes(x = EMP_GROUP, fill = SEX)) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = c("#66c2a5", "tan1"), name = "Gender") +
  labs(
    title = "Employment Status by Gender in 2024 April",
    subtitle = "More males are employed or unemployed, females are more likely to be out of the labor force",
    y = "Population",
    x = NULL,
    caption = "Source: IPUMS CPS Data"
  ) +
  theme_minimal()

```

The chart displays employment status by gender for 2024, highlighting three categories: Employed, Unemployed and Not in Labor Force (NILF). The majority of both genders are employed, with slightly more males than females. A larger share of females falls into the NILF category, suggesting possible factors such as caregiving responsibilities impacting their labor market participation. The unemployment rates are low for both genders, indicating stable employment for those seeking work. 


```{r}

```


```{r}

```


```{r}

```