---
title: "Assignment04"
author: "Xiang Xiang and Weijing Zhong"
format: html
---

### Data Source
[IPUMS CURRENT POPULATION SURVEY](https://cps.ipums.org/cps/index.shtml).

```{r}

library(ggplot2)
library(tidyverse)
library(dplyr)
library(ipumsr)

ddi <- read_ipums_ddi("cps_00003.xml")
data <- read_ipums_micro(ddi)

glimpse(data)

```





### Plot 1 Employment Status by Gender

```{r}

cps_2024 <- data %>%
  filter(YEAR == 2024, !EMPSTAT %in% c(0, 1))

cps_2024$SEX <- factor(cps_2024$SEX, levels = c(1, 2), labels = c("Male", "Female"))

cps_2024$EMP_GROUP <- ifelse(cps_2024$EMPSTAT %in% c(10, 12), "Employed",
                             ifelse(cps_2024$EMPSTAT %in% c(20, 21, 22), "Unemployed", "NILF"))

cps_2024$EMP_GROUP <- factor(cps_2024$EMP_GROUP, levels = c("Employed", "Unemployed", "NILF"))

ggplot(data = cps_2024, aes(x = EMP_GROUP, fill = SEX)) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = c("#66c2a5", "tan1"), name = "Gender") +
  labs(
    title = "Employment Status by Gender in 2024 April",
    subtitle = "More males are employed or unemployed, females are more likely to be out of the labor force",
    y = "Population",
    x = NULL,
    caption = "Source: IPUMS CPS Data"
  ) +
  theme_minimal()

```

The chart displays employment status by gender for 2024, highlighting three categories: Employed, Unemployed and Not in Labor Force (NILF). The majority of both genders are employed, with slightly more males than females. A larger share of females falls into the NILF category, suggesting possible factors such as caregiving responsibilities impacting their labor market participation. The unemployment rates are low for both genders, indicating stable employment for those seeking work. 




### Plot 2 Weekly Working Hours by Race

```{r}

data$RACE <- as.numeric(data$RACE)

data$RACE_GROUP <- case_when(
  data$RACE == 100 ~ "White",
  data$RACE == 200 ~ "Black",
  data$RACE == 300 ~ "American Indian/Alaskan Native",
  data$RACE == 650 ~ "Asian or Pacific Islander",
  data$RACE == 651 ~ "Asian or Pacific Islander",
  data$RACE == 652 ~ "Asian or Pacific Islander",
  data$RACE >= 801 & data$RACE <= 817 ~ "Two or More Races",
  TRUE ~ "Other"
)

data_clean <- data %>%
  filter(UHRSWORKT != 999 & UHRSWORKT != 997 & !is.na(UHRSWORKT)) %>%
  select(YEAR, RACE_GROUP, UHRSWORKT)

data_summary <- data_clean %>%
  group_by(YEAR, RACE_GROUP) %>%
  summarise(Average_Hours = mean(UHRSWORKT, na.rm = TRUE)) %>%
  ungroup()

ggplot(data = data_summary, aes(x = YEAR, y = Average_Hours, color = RACE_GROUP)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(
    title = "Average Weekly Hours Worked by Race Group from 2020 to 2024",
    subtitle = "All Race Groups Work Around 40 Hours Per Week",
    x = "Year",
    y = "Average Weekly Hours Worked",
    color = "Race Group",
    caption = "Source: IPUMS CPS Data"
  ) +
  scale_x_continuous(breaks = seq(2020, 2024, 1)) +
  theme_minimal()

```

The graph shows average weekly hours worked by racial groups from 2020 to 2024. Most groups maintain stable work hours around 38-40 per week, but the "Other" group shows a sharp rise in 2021 followed by a decline, dropping further in 2024. This trend suggests possible challenges in job access or economic participation for certain groups.


```{r}

```


```{r}

```